---
title: "Multiple Myeloma Pathway Analysis for spBART-Selected Risk Genes"
author: "Saurabh Bhandari"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 4
    code_folding: show
    theme: flatly
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = TRUE,
  warning = FALSE
)
```

# Overview

**Purpose:** Run functional enrichment analyses for the genes selected by spBART variable selection and persist the result objects to disk for reference.

**Scope:** This Rmd performs computation only. Formal tables and figures live in `Writeup Draft 5.tex`; this document does not reproduce them.

**Analyses performed:**

1. Gene Ontology (BP, MF, CC) over-representation with BH correction.
2. KEGG pathway over-representation.
3. MSigDB C2:CGP enrichment against MM-related signatures (MYELOMA, PLASMA_CELL, BORTEZOMIB, etc.).
4. MSigDB C7 enrichment against B-cell / plasma-cell immunologic signatures.
5. Over-representation against a literature-curated panel of ten MM-critical pathways.
6. Direct gene-pathway membership against the curated MM panel.

All outputs are saved as CSV and RDS in `../gene_set_analysis_results/`.

---

# Setup and Libraries

```{r install-and-load-packages, message=FALSE, warning=FALSE}
install_if_missing_cran <- function(pkg) {
  if (!require(pkg, character.only = TRUE, quietly = TRUE)) {
    message(sprintf("Installing %s from CRAN...", pkg))
    install.packages(pkg, dependencies = TRUE)
    library(pkg, character.only = TRUE)
  }
}

install_if_missing_bioc <- function(pkg) {
  if (!require(pkg, character.only = TRUE, quietly = TRUE)) {
    message(sprintf("Installing %s from Bioconductor...", pkg))
    if (!require("BiocManager", quietly = TRUE)) {
      install.packages("BiocManager")
    }
    BiocManager::install(pkg, update = FALSE, ask = FALSE)
    library(pkg, character.only = TRUE)
  }
}

cran_packages <- c("dplyr")
for (pkg in cran_packages) install_if_missing_cran(pkg)

bioc_packages <- c("clusterProfiler", "org.Hs.eg.db", "msigdbr",
                   "AnnotationDbi", "DOSE")
for (pkg in bioc_packages) install_if_missing_bioc(pkg)

message("\n=== All packages loaded successfully ===")
```

---

# Input: spBART-Selected Gene Set

```{r define-gene-vectors}
gene_vector_spBART <- c("IL1RAP", "ST5", "HERC6", "KL", "MYO1E", "ELK3", "CAPN2", "UBR4")
message(sprintf("spBART gene set: %d genes", length(gene_vector_spBART)))
```

---

# MM-Specific Gene Sets and Helper Functions

```{r helper-functions}
# ============================================================================
# MM-SPECIFIC GENE SETS FROM MSigDB
# ============================================================================

# MSigDB C2:CGP - Cancer Gene Perturbation signatures (includes MM studies)
msigdb_cgp <- msigdbr(species = "Homo sapiens", category = "C2", subcategory = "CGP") %>%
  dplyr::select(gs_name, entrez_gene)

mm_keywords <- c("MYELOMA", "PLASMA_CELL", "PLASMACYTOMA", "BORTEZOMIB",
                 "DEXAMETHASONE", "THALIDOMIDE", "LENALIDOMIDE", "VELCADE")
msigdb_mm_specific <- msigdb_cgp %>%
  filter(grepl(paste(mm_keywords, collapse = "|"), gs_name, ignore.case = TRUE))

# MSigDB C7 - Immunologic signatures (B-cell / plasma cell)
msigdb_immunologic <- msigdbr(species = "Homo sapiens", category = "C7", subcategory = "IMMUNESIGDB") %>%
  dplyr::select(gs_name, entrez_gene)

bcell_keywords <- c("BCELL", "B_CELL", "PLASMA", "MEMORY_B", "NAIVE_B", "ACTIVATED_B")
msigdb_bcell <- msigdb_immunologic %>%
  filter(grepl(paste(bcell_keywords, collapse = "|"), gs_name, ignore.case = TRUE))

message(sprintf("Loaded %d MM-specific gene sets from MSigDB C2:CGP",
                length(unique(msigdb_mm_specific$gs_name))))
message(sprintf("Loaded %d B-cell/plasma cell gene sets from MSigDB C7",
                length(unique(msigdb_bcell$gs_name))))

# ============================================================================
# CURATED MM PATHWAY GENE SETS (Literature-based)
# ============================================================================

symbol_to_entrez <- function(symbols) {
  mapped <- bitr(symbols, fromType = "SYMBOL", toType = "ENTREZID",
                 OrgDb = org.Hs.eg.db, drop = TRUE)
  return(mapped$ENTREZID)
}

mm_pathways_list <- list(
  "NFkB_Signaling" = c("NFKB1", "NFKB2", "RELA", "RELB", "REL", "NFKBIA", "NFKBIB",
                       "IKBKB", "IKBKG", "CHUK", "TRAF2", "TRAF3", "TRAF6", "NIK",
                       "MAP3K14", "BCL3", "CARD11", "MALT1", "BCL10"),
  "IL6_JAK_STAT" = c("IL6", "IL6R", "IL6ST", "JAK1", "JAK2", "JAK3", "TYK2",
                     "STAT1", "STAT3", "STAT5A", "STAT5B", "SOCS1", "SOCS3",
                     "PIAS1", "PIAS3", "SHP2", "PTPN11", "GP130"),
  "PI3K_AKT_mTOR" = c("PIK3CA", "PIK3CB", "PIK3CD", "PIK3CG", "PIK3R1", "PIK3R2",
                      "AKT1", "AKT2", "AKT3", "MTOR", "RPTOR", "RICTOR", "MLST8",
                      "PTEN", "PDK1", "PDPK1", "TSC1", "TSC2", "RHEB", "S6K1",
                      "RPS6KB1", "4EBP1", "EIF4EBP1"),
  "Cell_Cycle_MM" = c("CCND1", "CCND2", "CCND3", "CDK4", "CDK6", "RB1", "E2F1",
                      "E2F2", "E2F3", "CDKN1A", "CDKN1B", "CDKN2A", "CDKN2C",
                      "TP53", "MDM2", "MYC", "MYCN"),
  "Apoptosis_Survival" = c("BCL2", "BCL2L1", "BCL2L2", "MCL1", "BAX", "BAK1",
                           "BID", "BIM", "BCL2L11", "PUMA", "BBC3", "NOXA",
                           "PMAIP1", "XIAP", "BIRC5", "SURVIVIN"),
  "Bone_Remodeling" = c("TNFSF11", "RANKL", "TNFRSF11A", "RANK", "TNFRSF11B",
                        "OPG", "DKK1", "SOST", "WNT3A", "WNT5A", "LRP5", "LRP6",
                        "CTNNB1", "MIP1A", "CCL3", "IL3", "IL7", "HGF", "VEGFA"),
  "Proteasome_UPS" = c("PSMA1", "PSMB1", "PSMB5", "PSMD1", "PSMD2", "PSMD4",
                       "UBE2D1", "UBE2G1", "UBE2G2", "UBE2R1", "UBE3A", "UBE4B",
                       "USP7", "USP14", "UCHL5", "VCP", "NPLOC4", "UFD1"),
  "Plasma_Cell_Diff" = c("IRF4", "PRDM1", "BLIMP1", "XBP1", "PAX5", "BCL6",
                         "CIITA", "CD19", "CD38", "CD138", "SDC1", "SLAMF7",
                         "CS1", "BCMA", "TNFRSF17"),
  "DNA_Damage_Response" = c("TP53", "ATM", "ATR", "CHEK1", "CHEK2", "BRCA1",
                            "BRCA2", "RAD51", "PRKDC", "XRCC5", "XRCC6",
                            "DDB1", "DDB2", "XPC", "ERCC1", "MLH1", "MSH2"),
  "IL1_Signaling" = c("IL1A", "IL1B", "IL1R1", "IL1R2", "IL1RAP", "IL1RN",
                      "MYD88", "IRAK1", "IRAK2", "IRAK4", "TRAF6", "TAB1",
                      "TAB2", "TAK1", "MAP3K7", "NFKB1", "JNK", "MAPK8")
)

mm_pathways_t2g <- do.call(rbind, lapply(names(mm_pathways_list), function(pathway) {
  entrez_ids <- tryCatch({
    symbol_to_entrez(mm_pathways_list[[pathway]])
  }, error = function(e) character(0))

  if (length(entrez_ids) > 0) {
    data.frame(gs_name = pathway, entrez_gene = entrez_ids, stringsAsFactors = FALSE)
  } else {
    NULL
  }
}))

message(sprintf("Created %d curated MM pathway gene sets with %d total gene-pathway associations",
                length(unique(mm_pathways_t2g$gs_name)), nrow(mm_pathways_t2g)))

# ============================================================================
# Compute overlap between a gene vector and the curated MM pathway list
# ============================================================================
summarize_mm_pathway_overlap <- function(gene_vector, analysis_name) {
  message(sprintf("\n=== %s: MM Pathway Overlap Analysis ===", analysis_name))

  overlap_summary <- lapply(names(mm_pathways_list), function(pathway) {
    overlap <- intersect(gene_vector, mm_pathways_list[[pathway]])
    data.frame(
      Pathway = pathway,
      N_Pathway_Genes = length(mm_pathways_list[[pathway]]),
      N_Overlap = length(overlap),
      Overlap_Pct = round(100 * length(overlap) / length(mm_pathways_list[[pathway]]), 1),
      Genes = if (length(overlap) > 0) paste(overlap, collapse = ", ") else "-",
      stringsAsFactors = FALSE
    )
  })

  overlap_df <- do.call(rbind, overlap_summary)
  overlap_df <- overlap_df[order(-overlap_df$N_Overlap), ]

  pathways_with_overlap <- overlap_df$Pathway[overlap_df$N_Overlap > 0]
  message(sprintf("  Pathways with gene overlap: %d/%d", length(pathways_with_overlap),
                  length(mm_pathways_list)))
  if (length(pathways_with_overlap) > 0) {
    message(sprintf("  Pathways: %s", paste(pathways_with_overlap, collapse = ", ")))
  }

  return(overlap_df)
}

# ============================================================================
# Per-gene MM pathway membership
# ============================================================================
create_gene_mm_summary <- function(gene_vector) {
  gene_info <- lapply(gene_vector, function(gene) {
    pathways_containing <- names(mm_pathways_list)[sapply(mm_pathways_list, function(p) gene %in% p)]
    data.frame(
      Gene = gene,
      N_MM_Pathways = length(pathways_containing),
      MM_Pathways = if (length(pathways_containing) > 0) {
        paste(gsub("_", " ", pathways_containing), collapse = "; ")
      } else {
        "None (potentially novel)"
      },
      stringsAsFactors = FALSE
    )
  })

  gene_df <- do.call(rbind, gene_info)
  gene_df <- gene_df[order(-gene_df$N_MM_Pathways, gene_df$Gene), ]
  return(gene_df)
}

# ============================================================================
# Symbol -> Entrez for the spBART gene set
# ============================================================================
convert_genes_to_entrez <- function(gene_vector) {
  message(sprintf("\n========== Converting genes for spBART =========="))
  message(sprintf("Number of genes: %d", length(gene_vector)))

  gene_entrez <- bitr(gene_vector, fromType = "SYMBOL",
                      toType = c("ENTREZID", "ENSEMBL"), OrgDb = org.Hs.eg.db)
  message(sprintf("Mapped %d/%d genes to Entrez IDs", nrow(gene_entrez), length(gene_vector)))
  return(gene_entrez)
}
```

---

# Background Universe

```{r define-background-universe}
# Background = all human genes mappable to Entrez IDs via org.Hs.eg.db.
# If you have the actual ~19,100 profiled gene list from the 5hmC assay, load it
# here and convert to Entrez — that is the statistically correct background.
all_genes_symbols <- keys(org.Hs.eg.db, keytype = "SYMBOL")

background_entrez <- bitr(all_genes_symbols,
                          fromType = "SYMBOL",
                          toType = "ENTREZID",
                          OrgDb = org.Hs.eg.db)

message(sprintf("Background universe: %d genes with Entrez IDs", nrow(background_entrez)))
```

---

# Gene ID Conversion

```{r gene-conversion}
gene_entrez_spbart <- convert_genes_to_entrez(gene_vector_spBART)

output_dir <- "../gene_set_analysis_results"
if (!dir.exists(output_dir)) dir.create(output_dir, recursive = TRUE)
message(sprintf("Output directory: %s", normalizePath(output_dir)))
```

---

# GO Enrichment (BP, MF, CC)

```{r spbart-go-bp}
go_bp_spbart <- enrichGO(gene = gene_entrez_spbart$ENTREZID,
                         universe = background_entrez$ENTREZID,
                         OrgDb = org.Hs.eg.db,
                         ont = "BP",
                         pAdjustMethod = "BH",
                         pvalueCutoff = 0.1,
                         qvalueCutoff = 0.2,
                         readable = TRUE)
```

```{r spbart-go-mf}
go_mf_spbart <- enrichGO(gene = gene_entrez_spbart$ENTREZID,
                         universe = background_entrez$ENTREZID,
                         OrgDb = org.Hs.eg.db,
                         ont = "MF",
                         pAdjustMethod = "BH",
                         pvalueCutoff = 0.1,
                         qvalueCutoff = 0.2,
                         readable = TRUE)
```

```{r spbart-go-cc}
go_cc_spbart <- enrichGO(gene = gene_entrez_spbart$ENTREZID,
                         universe = background_entrez$ENTREZID,
                         OrgDb = org.Hs.eg.db,
                         ont = "CC",
                         pAdjustMethod = "BH",
                         pvalueCutoff = 0.1,
                         qvalueCutoff = 0.2,
                         readable = TRUE)
```

---

# KEGG Enrichment

```{r spbart-kegg}
kegg_spbart <- enrichKEGG(gene = gene_entrez_spbart$ENTREZID,
                          universe = background_entrez$ENTREZID,
                          organism = "hsa",
                          pAdjustMethod = "BH",
                          pvalueCutoff = 0.1,
                          qvalueCutoff = 0.2)

if (!is.null(kegg_spbart) && nrow(kegg_spbart) > 0) {
  kegg_spbart_readable <- setReadable(kegg_spbart, OrgDb = org.Hs.eg.db, keyType = "ENTREZID")
} else {
  kegg_spbart_readable <- kegg_spbart
}
```

---

# MSigDB MM-Specific (C2:CGP) Enrichment

```{r spbart-mm-specific}
if (nrow(msigdb_mm_specific) > 0) {
  mm_enrich_spbart <- enricher(gene = gene_entrez_spbart$ENTREZID,
                               TERM2GENE = msigdb_mm_specific,
                               pAdjustMethod = "BH",
                               pvalueCutoff = 0.1,
                               qvalueCutoff = 0.2)
} else {
  mm_enrich_spbart <- NULL
}
```

---

# MSigDB B-Cell / Plasma Cell (C7) Enrichment

```{r spbart-bcell}
if (nrow(msigdb_bcell) > 0) {
  bcell_enrich_spbart <- enricher(gene = gene_entrez_spbart$ENTREZID,
                                  TERM2GENE = msigdb_bcell,
                                  pAdjustMethod = "BH",
                                  pvalueCutoff = 0.1,
                                  qvalueCutoff = 0.2)
} else {
  bcell_enrich_spbart <- NULL
}
```

---

# Curated MM Pathway Enrichment

```{r spbart-mm-curated}
mm_curated_spbart <- enricher(gene = gene_entrez_spbart$ENTREZID,
                              TERM2GENE = mm_pathways_t2g,
                              pAdjustMethod = "BH",
                              pvalueCutoff = 0.2,
                              qvalueCutoff = 0.5)

if (!is.null(mm_curated_spbart) && nrow(mm_curated_spbart) > 0) {
  mm_curated_spbart@result$Description <- gsub("_", " ", mm_curated_spbart@result$Description)
}
```

---

# Direct MM Pathway Membership

```{r spbart-pathway-overlap}
spbart_overlap <- summarize_mm_pathway_overlap(gene_vector_spBART, "spBART")
spbart_gene_summary <- create_gene_mm_summary(gene_vector_spBART)

spbart_in_pathways <- spbart_gene_summary[spbart_gene_summary$N_MM_Pathways > 0, ]
spbart_novel <- spbart_gene_summary[spbart_gene_summary$N_MM_Pathways == 0, ]

pathway_summary <- data.frame(
  Pathway = names(mm_pathways_list),
  N_Genes = sapply(mm_pathways_list, function(p) sum(gene_vector_spBART %in% p))
)
pathway_summary <- pathway_summary[order(-pathway_summary$N_Genes), ]
pathway_summary$Pathway <- gsub("_", " ", pathway_summary$Pathway)
```

---

# Save Results

```{r save-results}
# GO
if (!is.null(go_bp_spbart) && nrow(go_bp_spbart) > 0) {
  write.csv(as.data.frame(go_bp_spbart),
            file.path(output_dir, "spbart_GO_BP.csv"), row.names = FALSE)
}
if (!is.null(go_mf_spbart) && nrow(go_mf_spbart) > 0) {
  write.csv(as.data.frame(go_mf_spbart),
            file.path(output_dir, "spbart_GO_MF.csv"), row.names = FALSE)
}
if (!is.null(go_cc_spbart) && nrow(go_cc_spbart) > 0) {
  write.csv(as.data.frame(go_cc_spbart),
            file.path(output_dir, "spbart_GO_CC.csv"), row.names = FALSE)
}

# KEGG
if (!is.null(kegg_spbart) && nrow(kegg_spbart) > 0) {
  write.csv(as.data.frame(kegg_spbart_readable),
            file.path(output_dir, "spbart_KEGG.csv"), row.names = FALSE)
}

# MSigDB
if (!is.null(mm_enrich_spbart) && nrow(mm_enrich_spbart) > 0) {
  write.csv(as.data.frame(mm_enrich_spbart),
            file.path(output_dir, "spbart_mm_specific_signatures.csv"), row.names = FALSE)
}
if (!is.null(bcell_enrich_spbart) && nrow(bcell_enrich_spbart) > 0) {
  write.csv(as.data.frame(bcell_enrich_spbart),
            file.path(output_dir, "spbart_bcell_signatures.csv"), row.names = FALSE)
}

# Curated MM pathway over-representation
if (!is.null(mm_curated_spbart) && nrow(mm_curated_spbart) > 0) {
  write.csv(as.data.frame(mm_curated_spbart),
            file.path(output_dir, "spbart_mm_curated_enrichment.csv"), row.names = FALSE)
}

# Direct pathway membership
write.csv(spbart_overlap,
          file.path(output_dir, "spbart_mm_pathway_overlap.csv"), row.names = FALSE)
write.csv(spbart_gene_summary,
          file.path(output_dir, "spbart_gene_mm_summary.csv"), row.names = FALSE)
write.csv(pathway_summary,
          file.path(output_dir, "spbart_mm_pathway_summary.csv"), row.names = FALSE)
write.csv(data.frame(Gene = gene_vector_spBART),
          file.path(output_dir, "spbart_genes.csv"), row.names = FALSE)
write.csv(spbart_novel,
          file.path(output_dir, "spbart_novel_genes.csv"), row.names = FALSE)

# Combined RDS
all_results <- list(
  gene_entrez_spbart = gene_entrez_spbart,
  background_universe = background_entrez,
  go_bp_spbart = go_bp_spbart,
  go_mf_spbart = go_mf_spbart,
  go_cc_spbart = go_cc_spbart,
  kegg_spbart = kegg_spbart,
  kegg_spbart_readable = kegg_spbart_readable,
  mm_enrich_spbart = mm_enrich_spbart,
  bcell_enrich_spbart = bcell_enrich_spbart,
  mm_curated_spbart = mm_curated_spbart,
  mm_pathway_overlap_spbart = spbart_overlap,
  mm_gene_summary_spbart = spbart_gene_summary,
  mm_pathway_summary = pathway_summary
)
saveRDS(all_results, file.path(output_dir, "spbart_enrichment_analysis_results.RDS"))

message(sprintf("\n=== Results saved to: %s ===", normalizePath(output_dir)))
```

---

# Session Info

```{r session-info}
sessionInfo()
```
